---
title: "Session 3: Biogeochemistry Data Tutorial"
author: Kaizad Patel
output: html_document
date: "2023-11-07"
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

**MONet Community Science Meeting 2023**

This RMarkdown document contains code and scripts for preliminary processing and data visualization of soil biogeochemistry data. 
These graphs were created for use in Session 3 (Biogeochemistry Data Tutorial) on Tuesday, November 7.


> **USING THIS TUTORIAL**  
Download the [bgc_data file](https://github.com/EMSL-MONet/CommSciMtg_Nov23/blob/main/Biogeochem/data/bgc_data.csv) and [bgc_analyses file](https://github.com/EMSL-MONet/CommSciMtg_Nov23/blob/main/Biogeochem/data/bgc_analyses.csv). 
Create a new R Script file and save it in the **same** folder as the data files. 
Make sure all these files are saved in the same folder, to minimize errors with file paths.


> This file is a Markdown report, which includes text as well as code chunks. You can copy the R code directly from these chunks and paste it into your R Script file. 


---

## Setup

### Load packages

You will need the `{tidyverse}` set of packages to run this tutorial. 
The `tidyverse` includes packages like `{dplyr}` and `{tidyr}` for data wrangling, and `{ggplot2}` for data visualization.

```{r packages}

library(tidyverse) # for general data wrangling and visualization
theme_set(theme_bw(base_size = 12)) # set default ggplot theme


library(ggcorrplot) # for correlation matrix plot
library(ggtern) # for soil texture triangle (ternary plot) 

```

You may also want to download additional packages to jazz up your figures, such as the `{soilpalettes}` package, which includes soil-themed color palettes.

```{r}
# Install `soilpalettes` using: 
# install.packages("devtools"); devtools::install_github("kaizadp/soilpalettes")

library(soilpalettes) # for color palettes
```


### Import files

We need two files for this tutorial. 
`bgc_data` includes biogeochemical variables, and also has site metadata such as latitude-longitude and biome type. `bgc_analyses` includes additional information about the analyses performed.

You will need to change the file path before using this tutorial. The current file path reflects the directory structure in the GitHub repository.  
If you downloaded the tutorial and the data file into the same folder, change the code chunk below to `bgc_data = read_csv("bgc_data.csv")` and ``bgc_analyses = read_csv("bgc_data.csv") %>% ...`


```{r import_files}
bgc_data = read_csv("Biogeochem/data/bgc_data.csv")
bgc_analyses = read_csv("Biogeochem/data/bgc_analyses.csv") %>% dplyr::select(analysis, analysis_type)
```


### Process the data

We will use the data in two forms - wideform (`bgc_data` is already in wideform, with each analyte as a separate column) and longform (see below).

```{r long}
data_long = 
  bgc_data %>% 
  pivot_longer(cols = -c(Site_Code, Location, Lat, Long, biome_name), 
               names_to = "analysis") %>% 
  left_join(bgc_analyses)

```


One additional processing step before we start the tutorial -- The `Location` column refers to the depth of the soil sample, "TOP" or "BTM".  
By default, R sorts its data alphabetically. Change the order here, before proceeding. 


```{r process}

bgc_data = 
  bgc_data %>% 
  mutate(Location = factor(Location, levels = c("TOP", "BTM")))

```


--- 

## BASIC EXPLORATION

Basic jitter plots to determine the range and distribution of each variable. Grouped by analysis type.

```{r jitters, fig.height=10, fig.width=12}
data_long %>% 
  ggplot(aes(x = analysis, y = value, color = Location))+
  geom_jitter(width = 0.3)+
  facet_wrap(~analysis_type, scales = "free")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


The samples were collected from sites spread across five biomes

```{r biome_numbers}
## Create summary of biome counts
biome_numbers = 
  data_long %>% 
  distinct(Site_Code, biome_name) %>% 
  group_by(biome_name) %>% 
  dplyr::summarise(n = n()) %>% 
  force()

biome_numbers %>% knitr::kable()
```

```{r biome_numbers_gg, eval=FALSE, echo=FALSE}

biome_numbers %>% 
  drop_na() %>% 
  ggplot(aes(x = 1, y = biome_name, fill = biome_name))+
  geom_point(aes(size = n), alpha = 1, shape = 21, color = "black", stroke = 1)+
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6)+
  geom_text(aes(label = biome_name, x = 1.25), 
            color = "black", hjust = 0, size = 6)+
  scale_fill_manual(values = soilpalettes::soil_palette("redox2", 5))+
  scale_radius(range = c(10, 32))+
  xlim(0, 5)+
  theme_void()+
  theme(legend.position = "none",
        panel.border = element_rect(fill = NA, color = "black"))
```



---

## CORRELATIONS

Use this function to set up the correlation plots for the entire dataset.  
We use the `ggcorrplot` package for this.

Creating a correlation matrix plot can be complicated. We have simplified this by making a custom function `fit_correlations_function`. This bundles all the processing/calculation steps into a neat function, which you can then run below.

```{r correlations_function}

  fit_correlations_function = function(dat, TITLE = ""){
    num = 
      dat %>%       
      dplyr::select(where(is.numeric)) %>%
      drop_na()
    
    num_clean = 
      num %>% 
      rownames_to_column("row") %>% 
      pivot_longer(-row) %>% 
      separate(name, sep = "_", into = c("name")) %>% 
      pivot_wider() %>% 
      dplyr::select(-row)
    
    
    m = cor(num_clean)
    p.mat <- ggcorrplot::cor_pmat(num_clean)
    
    ggcorrplot::ggcorrplot(m, type = "lower",
                           p.mat = p.mat,
                           outline.color = "black",
                           #   lab = TRUE, 
                           insig = "blank",
                           colors = c("#E46726", "white", "#6D9EC1"),
                           title = TITLE)
    
  }

```

Now, apply the `fit_correlations_function` function to the dataset to generate the correlation plot.

```{r correlation_matrix}
# this will generate a correlation matrix for the entire BGC dataset
  corr_all = fit_correlations_function(bgc_data %>% dplyr::select(-Lat, -Long, -GWC))

# However, there are a lot of variables here. 
# So, for an easier plot, subset select variables and then re-run the correlation matrix.

data_subset = bgc_data %>% 
  dplyr::select(Ca, Mg, CEC, 
                Clay, SO4S, NH4N,
                TC, TN, Location)

fit_correlations_function(data_subset %>% filter(Location == "TOP"))
```

**Individual Correlations**

Here is simple code to visualize correlations among variables. You can also do this in the [Shiny App](https://shinyproxy.emsl.pnnl.gov/app/1000soils).

```{r correlations_individual}

bgc_data %>% 
  ggplot(aes(x = TC, y = TN))+
  geom_point()+
  geom_smooth(method = "lm", se = F)+
  labs(x = "Total C (%)",
       y = "Total N (%)")


bgc_data %>% 
  ggplot(aes(x = Clay, y = Sand))+
  geom_point()+
  geom_smooth(method = "lm", se = F)+
  labs(x = "Clay (%)",
       y = "Sand (%)")


bgc_data %>% 
  ggplot(aes(y = CEC, x = Clay))+
  geom_point()+
  geom_smooth(method = "lm", se = F)+
  labs(y = "Cation Exchange Capacity (meq/100g)",
       x = "Clay (%)")


bgc_data %>% 
  ggplot(aes(x = Ca, y = Bases))+
  geom_point()+
  geom_smooth(method = "lm", se = F)+
  labs(y = "Total bases (meq/100g)",
       x = "Calcium (meq/100g)")



bgc_data %>% 
  ggplot(aes(x = TC, y = WEOC))+
  geom_point()+
  geom_smooth(method = "lm", se = F)+
  labs(x = "Total carbon (%)",
       y = "Water-extractable organic carbon (mg/g)")

```

--- 

## JITTER PLOTS

These are "jittered" scatter plots showing distribution of the variables by depth or by biome.  
Also included is code to calculate the analysis of variance (ANOVA)


### 1. total carbon (TC)

TC ANOVA

```{r}

lm((TC) ~ Location * biome_name, data = bgc_data) %>% car::Anova()

```


plot TC by depth

```{r}

bgc_data %>% 
  ggplot(aes(x = Location, y = TC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3, outlier.colour = NA)+
  geom_jitter(size = 3, width = 0.3)

```

plot TC by biome

```{r, fig.height=5, fig.width=9}
bgc_data %>% 
  ggplot(aes(x = biome_name, y = TC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3)+
  geom_point(size = 3, position = position_dodge(width = 0.5))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  labs(x = "")
```


### 2. CEC

CEC ANOVA

```{r}

lm((CEC) ~ Location * biome_name, data = bgc_data) %>% car::Anova()

```

plot CEC by depth

```{r}

bgc_data %>% 
  ggplot(aes(x = Location, y = CEC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3, outlier.colour = NA)+
  geom_jitter(size = 3, width = 0.3)+
  labs(x = "")

```

plot CEC by biome

```{r, fig.height=5, fig.width=9}
bgc_data %>% 
  ggplot(aes(x = biome_name, y = CEC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3)+
  geom_point(size = 3, position = position_dodge(width = 0.5))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  labs(x = "")
```


### 3. Clay

Clay ANOVA

```{r}

lm((Clay) ~ Location * biome_name, data = bgc_data) %>% car::Anova()

```

plot Clay by depth

```{r}

bgc_data %>% 
  ggplot(aes(x = Location, y = Clay, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3, outlier.colour = NA)+
  geom_jitter(size = 3, width = 0.3)+
  labs(x = "")

```

plot Clay by biome

```{r, fig.height=5, fig.width=9}
bgc_data %>% 
  ggplot(aes(x = biome_name, y = Clay, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3)+
  geom_point(size = 3, position = position_dodge(width = 0.5))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  labs(x = "")
```



### 4. WEOC

WEOC ANOVA

```{r}

lm((WEOC) ~ Location * biome_name, data = bgc_data) %>% car::Anova()

```

plot WEOC by depth

```{r}

bgc_data %>% 
  ggplot(aes(x = Location, y = WEOC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3, outlier.colour = NA)+
  geom_jitter(size = 3, width = 0.3)+
  labs(x = "")

```

plot WEOC by biome

```{r, fig.height=5, fig.width=9}
bgc_data %>% 
  ggplot(aes(x = biome_name, y = WEOC, color = Location))+
  geom_boxplot(position = position_dodge(width = 0.5), width = 0.3)+
  geom_point(size = 3, position = position_dodge(width = 0.5))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  labs(x = "")
```


---

## MAPS

The `{ggplot2}` package has functionality to easily plot maps (without downloading any other files or packages!).

We will plot our data over a map of the USA.  
To do so, first load the `states_map` file, which contains the data for the state outlines.  
Then, plot the state boundaries using the `geom_polygon` function.

```{r}
states_map <- ggplot2::map_data("state")

ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)
```

Now, to plot our data over this map.  
We add the `geom_point` layer over the base polygon layer of the states.

```{r}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data, 
             aes(x = Long, y = Lat, 
                 fill = TC, group = 1), 
             size = 6, shape = 21, color = "black")
```


The default `{ggplot2}` color palette may be difficult to interpret.  
As an alternative, we use the `{soilpalettes}` package.  

```{r, fig.height=7, fig.width=11}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data %>% filter(Location == "TOP"), 
             aes(x = Long, y = Lat, color = TC, group = 1), 
             size = 6)+
  scale_color_gradientn(colors = soil_palette("redox2", 5))+
  labs(color = "TC, %")+
  theme_void(base_size = 16)+
  theme(legend.position = "bottom")
```

```{r, fig.height=7, fig.width=11}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data %>% filter(Location == "TOP"), 
             aes(x = Long, y = Lat, color = WEOC, group = 1), 
             size = 6)+
  scale_color_gradientn(colors = soil_palette("redox2", 5))+
  labs(color = "WEOC, mg/g")+
  theme_void(base_size = 16)+
  theme(legend.position = "bottom")
```


```{r, fig.height=7, fig.width=11}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data %>% filter(Location == "TOP"), 
             aes(x = Long, y = Lat, color = CEC, group = 1), 
             size = 6)+
  scale_color_gradientn(colors = soil_palette("redox2", 5))+
  labs(color = "CEC, meq/100g")+
  theme_void(base_size = 16)+
  theme(legend.position = "bottom")
```


```{r, fig.height=7, fig.width=11}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data %>% filter(Location == "TOP"), 
             aes(x = Long, y = Lat, color = Clay, group = 1), 
             size = 6)+
  scale_color_gradientn(colors = soil_palette("redox2", 5))+
  labs(color = "Clay, %")+
  theme_void(base_size = 16)+
  theme(legend.position = "bottom")
```

Another map example, showing the different biomes

```{r, fig.height=7, fig.width=17}
ggplot(states_map, aes(long, lat, group = group))+
  geom_polygon(color = "grey", fill = NA)+
  geom_point(data = bgc_data, 
             aes(x = Long, y = Lat, 
                 color = biome_name, group = 1), 
             size = 6)+
  scale_color_manual(values = soil_palette("redox2", 5))+
  theme_void(base_size = 16)+
  labs(fill = "")
  
```


---

## TEXTURE

We plot % sand-silt-clay on a texture triangle (ternary plot) to determine the soil texture. We use the `ggtern` package for this. 

You first plot the base ternary triangle. You will need to load the `USDA` dataset from `ggtern`.

```{r}
data(USDA)

 texture_base = 
  ggtern() +
    geom_polygon(data = USDA, 
                 aes(x = Sand, y = Clay, z = Silt, group = Label),
                 fill = NA, size = 0.3, alpha = 0.5, color = "grey30")+
    theme_bw()+
    theme_showarrows()+
    theme_hidetitles()+
    theme_clockwise()

texture_base
```

then overlay your data points

```{r}

texture_base +
  geom_point(data = bgc_data,
             aes(x = Sand, y = Clay, z = Silt),
             size = 4, color = "black")

```

Now, color the data points by CEC
```{r}
texture_cec = 
  texture_base +
  geom_point(data = bgc_data %>% filter(!is.na(CEC)),
             aes(x = Sand, y = Clay, z = Silt, color = CEC),
             size = 4, )+
  scale_color_gradientn(colors = soil_palette("redox2", 5))+
  labs(fill = "CEC (meq/100g)")


```

Now, we need to add the texture labels.  
Create a new dataframe from the `USDA` dataset (included with `ggtern`). Then add that to the previous plot.

```{r, fig.height=8, fig.width=8}
  USDA_text <- 
    USDA  %>% 
    group_by(Label) %>%
    summarise_if(is.numeric, mean, na.rm = TRUE) %>%
    ungroup()
  
  #textures_names<-
  texture_cec +
    geom_text(data = USDA_text,
              aes(x = Sand, y = Clay, z = Silt, label = Label),
              size = 4, color = "black")
  
```

