---
title: "MONet FT-ICR-MS Hands -on Session"
output: html_document
---
This script will read in the molecular formula assignments from CoreMS and plot common figures.
The data is from the 1000 Soils Pilot project and has been:

* Formula assigned, filtered and aligned in Formularity format ('icr.reps.combined_v1.csv')
* Summarized, averaged and merged with soil chemistry data ('Summary_Merged_ZenodoShinyAp.csv')

The workflow here is split into two parts
*Part 1* is a qualitative examination of peak-level data, where *Part 2* aggregates properties by sample and then compares to additional data in a more quantitative way.
This is the 'extra-credit' part of the task\


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```



# **Part 1: Qualitative Data Exploration at the Formula Level**

## Loading Libraries
```{r}
library(ggplot2)
library(dplyr)
library(reshape2)
```



## Loading Data
```{r}
longform = read.csv("Longform_Formula_Level.csv")
summaryfile = read.csv("Merged_SummaryFile_ZenodoShinyAp.csv")

init_column = c(
  'Molecular.Formula', 'C', 'H', 'O', 'N', #'C13', 
                'S', 'P', #'Na',
           'El_comp', 'Class',#'NeutralMass',
                'Error_ppm', 
                'Candidates', 'H.C','O.C'
)

El_comp_order = c("CHO", "CHON")

# List and sort all of the unique Samples
sample_list = unique(longform$Sample)
sample_list = sort(sample_list)


# List out Biomes
biomes = c("Desert_Shrubland", "Medit_Woodlands", "Temp_Conif_Forests", "Temp_Forests", "Temp_Grasslands")
```

## Scatter Plot O/C and H/C 
Sample: ANZA_BTM\
Coloring: Class\
```{r}
data = filter(longform, Sample =="ANZA_BTM")
ggplot(data = data, aes(x = O/C, y = H/C, color = Class)) + geom_point() + labs(title = "Sample: ANZA_BTM")
```


## Scatter Plot C and DBE
Sample: ANZA_BTM\
Coloring: Class\
```{r}
data = filter(longform, Sample =="ANZA_BTM")
ggplot(data = data, aes(x = C, y = DBE, color = Class)) + geom_point() + labs(title = "Sample: ANZA_BTM")
```


## Scatter Plot C and Error_ppm 
Sample: ANZA_BTM\
Coloring: Class\
```{r}
data = filter(longform, Sample =="ANZA_BTM")
ggplot(data = data, aes(x = C, y = Error_ppm, color = Class)) + geom_point() + labs(title = "Sample: ANZA_BTM")
```


## Scatter Plot O/C and H/C 
Site: BLAN\
Coloring: Class\
```{r}
data = filter(longform, Site =="BLAN")
ggplot(data = data, aes(x = O/C, y = H/C, color = Class)) + geom_point() + labs(title = "Site: BLAN")
```



## Scatter Plot C and DBE
Site: BLAN\
Coloring: Class\
```{r}
data = filter(longform, Site =="BLAN")
ggplot(data = data, aes(x = C, y = DBE, color = Class)) + geom_point() + labs(title = "Site: BLAN")
```



## Scatter Plot C and Error_ppm 
Site BLAN\
Coloring: Class\
```{r}
data = filter(longform, Site =="BLAN")
ggplot(data = data, aes(x = C, y = Error_ppm, color = Class)) + geom_point() + labs(title = "Site: BLAN")
```



## Scatter Plot O/C and H/C 
Depth: TOP\
Coloring: Class\
```{r}
data = filter(longform, Depth =="TOP")
ggplot(data = data, aes(x = O/C, y = H/C, color = Class)) + geom_point() + labs(title = "Depth: TOP")
```


## Scatter Plot C and DBE 
Depth: TOP\
Coloring: Class\
```{r}
data = filter(longform, Depth =="TOP")
ggplot(data = data, aes(x = C, y = DBE, color = Class)) + geom_point() + labs(title = "Depth: TOP")
```


## Scatter Plot C and Error_ppm
Depth: TOP\
Coloring: Class\
```{r}
data = filter(longform, Depth =="TOP")
ggplot(data = data, aes(x = C, y = Error_ppm, color = Class)) + geom_point() + labs(title = "Depth: TOP")
```



## Scatter Plot O/C and H/C 
BiomeType: Temp_Forests\
Coloring by Class\
```{r}
data = filter(longform, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = O/C, y = H/C, color = Depth)) + geom_point() + labs(title = "BiomeType: Temp_Forests")
```


## Scatter Plot C and DBE 
Coloring: Class\
BiomeType: Temp_Forests\
```{r}
data = filter(longform, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = C, y = DBE, color = Class)) + geom_point() + labs(title = "BiomeType: Temp_Forests")
```



## Scatter Plot C and Errpr_ppm 
Coloring: Class\
BiomeType: Temp_Forests\
```{r}
data = filter(longform, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = C, y = Error_ppm, color = Class)) + geom_point() + labs(title = "BiomeType: Temp_Forests")
```


## Panel of Scatter Plots O/C and H/C '
BiomeType: Temp_Forests\
Coloring: Class\
Panels by Site\
```{r}
data = filter(longform, BiomeType == "Temp_Grasslands")
ggplot(data = data, aes(x = O/C, y = H/C, color = El_comp)) + geom_point() + labs(title = "BiomeType: Temp_Grasslands") + facet_wrap(~Site, ncol=5)
```




# **Part 2: Quantitative Comparison of Summed Formulas and Averaged Calculated Properties**

## Creating the data
```{R}
data = subset(summaryfile, select = c("ID", "BiomeType", "Site", "Depth", "CHO", "CHON"))
columns = c("ID", "BiomeType", "Site", "Depth")
value_vars = c("CHO", "CHON")
melt_data = melt(data, id=columns, variable.name = "El_comp", value.name = "El_comp_percent")

classes = c("Lignin", "Lipid", "ConHC", "Carb", "AminoSugar", "Tannin", "Protein", "UnsatHC", "Other")
class_data <- subset(summaryfile, select = c("ID", "BiomeType", "Site", "Depth" , "Assignments", classes))
melt_Classes = melt(class_data, id=c(columns, "Assignments"), variable.name = "Class", value.name = "Class_Value")

Biomes = c("Desert_Shrubland", "Temp_Forests", "Temp_Conif_Forests", "Temp_Grasslands", "Medit_Woodlands")
```



## Barplot for Site and Assignments
Site: RMNP\
Filling: Depth\
```{R}
data = filter(summaryfile, Site == "RMNP")
ggplot(data = data, aes(x = Site, y = Assignments, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge()) + scale_y_continuous(breaks = c(0,2000,4000,6000,8000)) + labs(fill="Depth")
```


## Assignmnets for Site and Assignments
BiomeType: Temp_Grasslands\
Filling: Depth\
```{R}
data = filter(summaryfile, BiomeType == "Temp_Grasslands")
ggplot(data = data, aes(x = Site, y = Assignments, fill= Depth)) + geom_bar(stat = "identity", position = position_dodge()) + scale_y_continuous(breaks = c(0,2000,4000,6000,8000)) + labs(fill="Depth")  + theme(axis.text.x = element_text(angle = 90))
```


## Barplot for Site and NOSC_Mean
Site: CFS1\
Filling: Depth\
```{R}
data = filter(summaryfile, Site == "CFS1")
ggplot(data = data, aes(x = Site, y = NOSC_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth")
```


## Barplot for Site and NOSC_Mean
Filling: Depth\
```{R}
ggplot(data = summaryfile, aes(x = Site, y = NOSC_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```


## Barplot for Site and NOSC_Mean
BiomeType: Temp_Forests\
Filling: Depth\
```{R}
data = filter(summaryfile, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = Site, y = NOSC_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```



## Barplot for Site and AImod_Mean
Site: CFS1\
Filling: Depth\
```{R}
data = filter(summaryfile, Site == "CFS1")
ggplot(data = data, aes(x = Site, y = AImod_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth")
```



## Barplot for Site and AI_Mean
Site: CFS1\
Filling: Depth\
```{R}
data = filter(summaryfile, Site == "CFS1")
ggplot(data = data, aes(x = Site, y = AI_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth")
```


## Barplot for Site and AI_Mean
BiomeType: Temp_Forests\
Filling: Depth\
```{R}
data = filter(summaryfile, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = Site, y = AI_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```



## Barplot for Site and DBE_Mean
Site: CFS1\
Filling: Depth\
```{R}
data = filter(summaryfile, Site == "CFS1")
ggplot(data = data, aes(x = Site, y = DBE_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth")
```


## Barplot for Site and DBE_Mean
Depth: TOP\
Filling: Depth\
```{R}
data = filter(summaryfile, Depth == "TOP")
ggplot(data = data, aes(x = Site, y = DBE_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```



## Barplot for Site and DBE_Mean
BiomeType: Temp_Forests\
Filling: Depth\
```{R}
data = filter(summaryfile, BiomeType == "Temp_Forests")
ggplot(data = data, aes(x = Site, y = DBE_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_bar(stat = "identity", position = position_dodge())  + labs(fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```


## Boxplot for BiomeType and NOSC_Mean
BiomeType: All biomes\
Filling: Depth\
```{R}
data = filter(summaryfile, BiomeType %in% Biomes)
ggplot(data = data, aes(x = factor(BiomeType, levels = Biomes), y = NOSC_Mean, fill= factor(Depth, levels = c("TOP", "BTM")))) + geom_boxplot() + labs(x="BiomeType",title = "BiomeType",fill="Depth") + theme(axis.text.x = element_text(angle = 90))
```




## Boxplot for Assignments and Mean_Error_PPM
Filling: Depth\
```{r echo=FALSE}
ggplot(data = summaryfile, aes(x = Assignments, y = Mean_Error_PPM, fill = Depth)) + geom_boxplot() + labs(title = "Site: All Sites") 

```




## Barplot for Site and Assignments
BiomeType: Temp_Forests\
Filling: Depth\
```{r echo=FALSE}
Class1 = melt_Classes %>% dplyr::group_by(Class, Site) %>% dplyr::summarise(Mean_Fraction=mean(Class_Value), SE=sd(Class_Value)/sqrt(length(Class_Value))) %>% filter(Site == "CFS1")

ggplot(data = Class1, aes(x = Class, y = Mean_Fraction, fill = Class)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90)) + geom_errorbar(aes(x = Class, ymin= Mean_Fraction - SE, ymax = Mean_Fraction + SE))

```



## Barplot for Class and Mean_Fraction
BiomeType: Temp_Forests\
Filling: Class\
```{r echo=FALSE}
Class1 = melt_Classes %>% dplyr::group_by(Class, Site, BiomeType, Assignments) %>% dplyr::summarise(Mean_Fraction=mean(Class_Value), SE=sd(Class_Value)/sqrt(length(Class_Value)), Assignments=Assignments) %>% filter(BiomeType == "Temp_Conif_Forests")

ggplot(data = Class1, aes(x = Class, y = Mean_Fraction, fill = Class)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Site) + geom_errorbar(aes(x = Class, ymin = Mean_Fraction - SE, ymax = Mean_Fraction + SE))

```




## Barplot for Class and Mean_Fraction
BiomeType: Temp_Forests\
Filling: Depth\
```{r echo=FALSE}
Class1 = melt_Classes %>% dplyr::group_by(Class, Site, Depth) %>% dplyr::summarise(Mean_Fraction=mean(Class_Value), SE=sd(Class_Value)/sqrt(length(Class_Value))) %>% filter(Site == "CFS1")

ggplot(data = Class1, aes(x = Class, y = Mean_Fraction, fill = Class)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~factor(Depth, levels = c("TOP", "BTM"))) 

```




## Barplot for BiomeType and Mean_Fraction
BiomeType: Temp_Forests\
Filling: Depth\
```{r echo=FALSE}
El_comp_data = melt_data %>% dplyr::group_by(El_comp, BiomeType) %>% dplyr::summarise(Mean_Fraction=mean(El_comp_percent), SE=sd(El_comp_percent)/sqrt(length(El_comp_percent))) %>% dplyr::filter(BiomeType != "None") %>% filter( BiomeType %in% biomes)

ggplot(data = El_comp_data, aes(x = BiomeType, y = Mean_Fraction, fill = El_comp)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90))

```




## Barplot for Site and Mean_Fraction
BiomeType: Temp_Forests\
Filling: Depth\
```{r echo=FALSE}
El_comp_data = melt_data %>% dplyr::group_by(El_comp, BiomeType, Depth, Site) %>% dplyr::summarise(Mean_Fraction=mean(El_comp_percent), SE=sd(El_comp_percent)/sqrt(length(El_comp_percent))) %>% dplyr::filter(BiomeType != "None") %>% filter( BiomeType %in% biomes)

data = filter(El_comp_data, Site == "CFS1")
ggplot(data = data, aes(x = Site, y = Mean_Fraction, fill = El_comp)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Depth)
```




## Barplot for Site and Mean_Fraction
BiomeType: Temp_Forests\
Filling: Depth\
```{r echo=FALSE}
El_comp_data = melt_data %>% dplyr::group_by(El_comp, BiomeType, Depth, Site) %>% dplyr::summarise(Mean_Fraction=mean(El_comp_percent), SE=sd(El_comp_percent)/sqrt(length(El_comp_percent))) %>% dplyr::filter(BiomeType != "None") %>% filter( BiomeType %in% biomes)

data = filter(El_comp_data, Site == "CFS1" & Depth == "TOP")
ggplot(data = data, aes(x = Site, y = Mean_Fraction, fill = El_comp)) + geom_bar(stat = "identity", position = position_dodge())  + labs(title = "All Biomes") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Depth)
```




## Boxplot for BiomeType and Mean_Error_PPM
Depth: TOP\
BiomeType: All biomes\
Filling: BiomeType\
```{r echo=FALSE}
data = filter(summaryfile, Depth == "TOP" & BiomeType %in% biomes)
ggplot(data = data, aes(x = BiomeType ,y=Mean_Error_PPM, fill=BiomeType)) + geom_boxplot() + labs(title = "Depth: TOP")  + theme(axis.text.x = element_text(angle = 90))

```